Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-15 article.Rmd
Finish revising introduction

It is important to understand our method's limitations as well as its advantages. The approach of inferring gender from personal names is still a blunt way to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept, including the fluid relationship between biological sex and individual identity [Butler 1990, Butler 1993]. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, or what the categories “male” or “female” actually mean in any given historical and social context [Crenshaw 1991, McCall 2005]. Perhaps most importantly, the tool itself does not reflect how an individual person self-identifies and how gender shapes their lived experience. The question of gender identity is especially relevant for our method, given its reliance on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose their own vision of a gender binary onto their subjects in ways that are deeply problematic. This gender binary reflects existing power structures more than the lived identities of individuals. The state's gender categorization exerts a tremendous definitional sway over societies and the people who comprise them, most notably in its systematic exclusion of transgender identities. [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014].

Researchers who use this method need to be fully aware of the ways it replicates a problematic, state-defined gender binary. The onus is on the user to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals who stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

>>>>

Researchers who use this method need to be fully aware of its limitations, in particular its dependency on a state-defined gender binary. Inferring gender from personal names is a blunt tool to study a complex subject. Gender theorists and feminist scholars have spent decades unpacking the full range of meanings behind gender as an analytical concept, including the fluid relationship between biological sex and gender identity [Butler 1990, Butler 1993]. On its own, our method has little to say about crucial topics such as the intersectionality of gender with race, class, and power, or what the categories "male" or "female" mean in any given historical and social context [Crenshaw 1991, McCall 2005]. The tool cannot reflect how an individual person self-identifies and how gender shapes their lived experience, since it relies on large datasets created by state agencies such as the U.S. Census Bureau. Much like the state’s definition of racial categories, governments impose a gender binary in their data collection that reflects existing power structures more than the lived identities of individuals. [Brubaker and Cooper 2000, 14-17, Pascoe 2009, Canaday 2011, Landrum 2014]. When using the predictions provided by this method, users must judge for themselves how well they reflect the social and cultural practices of the time and place under study.

Imperfect as it admittedly is, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals who stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.
